---
title: "Croarkin-Brandon_Project1"
author: "Brandon Croarkin"
date: "July 29, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Project 1 - HO2 Analysis

###Part 1: HO2 Data Preperation and Exploration

####Problem

```{r}
# Read in data package EIAdata
HO2 <- read.csv("Data/nyhh02.csv", header = T, 
    stringsAsFactors = F)
# stringsAsFactors sets dates as
# character type
head(HO2)
HO2 <- na.omit(HO2)  ## to clean up any missing data
# use na.approx() as well
str(HO2)  # review the structure of the data so far
```

####Questions

1. What is the nature of HO2 returns? We want to reflect the ups and downs of price movements, something of prime interest to management. First, we calculate percentage changes as log returns. Our interest is in the ups and downs. To look at that we use if and else statements to define a new column called  direction. We will build a data frame to house this analysis.

```{r}
# Construct expanded data frame
return <- as.numeric(diff(log(HO2$DHOILNYH))) * 
    100  # Euler 
size <- as.numeric(abs(return))  # size is indicator of volatility
direction <- ifelse(return > 0, "up", 
    ifelse(return < 0, "down", "same"))  # another indicator of volatility
# =if(return > 0, 'up', if(return <
# 0, 'down', 'same'))
date <- as.Date(HO2$DATE[-1], "%m/%d/%Y")  # length of DATE is length of return +1: omit 1st observation
price <- as.numeric(HO2$DHOILNYH[-1])  # length of DHOILNYH is length of return +1: omit first observation
HO2.df <- na.omit(data.frame(date = date, 
    price = price, return = return, size = size, 
    direction = direction))  # clean up data frame by omitting NAs
str(HO2.df)
```

We can plot with the ggplot2 package. In the ggplot statements we use aes, "aesthetics", to pick x (horizontal) and y (vertical) axes. Use group =1 to ensure that all data is plotted. The added (+)  geom_line is the geometrical method that builds the line plot.

```{r}
library(ggplot2)
p <- ggplot(HO2.df, aes(x = date, y = return, 
    group = 1)) + geom_line(colour = "blue")
p
```

Let's try a bar graph of the absolute value of price rates. We use  geom_bar to build this picture.

```{r}
p <- ggplot(HO2.df, aes(x = date, y = size, 
    group = 1)) + geom_bar(stat = "identity", 
    colour = "green")
p
```

Now let's build an overlay of return on size.

```{r}
p <- ggplot(HO2.df, aes(date, size)) + 
    geom_bar(stat = "identity", colour = "darkorange") + 
    geom_line(data = HO2.df, aes(date, 
        return), colour = "blue")
p
```

2. Let's dig deeper and compute mean, standard deviation, etc. Load the data_moments() function. Run the function using the  HO2.df$return subset of the data and write a knitr::kable() report.

```{r}
# Load the data_moments() function
# data_moments function INPUTS:
# vector OUTPUTS: list of scalars
# (mean, sd, median, skewness,
# kurtosis)
data_moments <- function(data) {
    library(moments)
    mean.r <- mean(data)
    sd.r <- sd(data)
    median.r <- median(data)
    skewness.r <- skewness(data)
    kurtosis.r <- kurtosis(data)
    result <- data.frame(mean = mean.r, 
        std_dev = sd.r, median = median.r, 
        skewness = skewness.r, kurtosis = kurtosis.r)
    return(result)
}
# Run data_moments()
answer <- data_moments(HO2.df$return)
# Build pretty table
answer <- round(answer, 4)
knitr::kable(answer)
```

3. Let's pivot size and return on direction. What is the average and range of returns by direction? How often might we view positive or negative movements in HO2?

```{r}
# Counting
table(HO2.df$return < 0)  # one way
table(HO2.df$return > 0)
table(HO2.df$direction)  # this counts 0 returns as negative
table(HO2.df$return == 0)
# Pivoting
library(dplyr)
## 1: filter to those houses with
## fairly high prices pivot.table <-
## filter(HO2.df, size >
## 0.5*max(size)) 2: set up data frame
## for by-group processing
pivot.table <- group_by(HO2.df, direction)
## 3: calculate the summary metrics
options(dplyr.width = Inf)  ## to display all columns
HO2.count <- length(HO2.df$return)
pivot.table <- summarise(pivot.table, 
    return.avg = round(mean(return), 
        4), return.sd = round(sd(return), 
        4), quantile.5 = round(quantile(return, 
        0.05), 4), quantile.95 = round(quantile(return, 
        0.95), 4), percent = round((length(return)/HO2.count) * 
        100, 2))
# Build visual
knitr::kable(pivot.table, digits = 2)
# Here is how we can produce a LaTeX
# formatted and rendered table
library(xtable)
HO2.caption <- "Heating Oil No. 2: 1986-2016"
print(xtable(t(pivot.table), digits = 2, 
    caption = HO2.caption, align = rep("r", 
        4), table.placement = "V"))
print(xtable(answer), digits = 2)
```

###Part 2

We will use the data from Part 1 to investigate the distribution of returns we generated. This will entail fitting the data to some parametric distributions as well as

####Problem

We want to further characterize the distribution of up and down movements visually. Also we would like to repeat the analysis periodically for inclusion in management reports.

####Questions

1. How can we show the differences in the shape of ups and downs in HO2, especially given our tolerance for risk? We can use the HO2.df data frame with ggplot2 and the cumulative relative frequency function stat_ecdf to begin to understand this data.

```{r}
HO2.tol.pct <- 0.95
HO2.tol <- quantile(HO2.df$return, HO2.tol.pct)
HO2.tol.label <- paste("Tolerable Rate = ", 
    round(HO2.tol, 2), sep = "")
ggplot(HO2.df, aes(return, fill = direction)) + 
    stat_ecdf(colour = "blue", size = 0.75) + 
    geom_vline(xintercept = HO2.tol, 
        colour = "red", size = 1.5) + 
    annotate("text", x = HO2.tol + 5, 
        y = 0.75, label = HO2.tol.label, 
        colour = "darkred")
```

2. How can we regularly, and reliably, analyze HO2 price movements? For this requirement, let's write a function similar to data_moments. Name this new function HO2_movement().

```{r}
## HO2_movement(file, caption) input:
## HO2 csv file from /data directory
## output: result for input to kable
## in $table and xtable in $xtable;
## data frame for plotting and further
## analysis in $df.  Example: HO2.data
## <- HO2_movement(file =
## 'data/nyhh02.csv', caption = 'HO2
## NYH')
HO2_movement <- function(file = "data/nyhh02.csv", 
    caption = "Heating Oil No. 2: 1986-2016") {
    # Read file and deposit into variable
    HO2 <- read.csv(file, header = T, 
        stringsAsFactors = F)
    # stringsAsFactors sets dates as
    # character type
    HO2 <- na.omit(HO2)  ## to clean up any missing data
    # Construct expanded data frame
    return <- as.numeric(diff(log(HO2$DHOILNYH))) * 
        100
    size <- as.numeric(abs(return))  # size is indicator of volatility
    direction <- ifelse(return > 0, "up", 
        ifelse(return < 0, "down", "same"))  # another indicator of volatility
    date <- as.Date(HO2$DATE[-1], "%m/%d/%Y")  # length of DATE is length of return +1: omit 1st observation
    price <- as.numeric(HO2$DHOILNYH[-1])  # length of DHOILNYH is length of return +1: omit first observation
    HO2.df <- na.omit(data.frame(date = date, 
        price = price, return = return, 
        size = size, direction = direction))  # clean up data frame by omitting NAs
    require(dplyr)
    ## 1: filter if necessary pivot.table
    ## <- filter(HO2.df, size >
    ## 0.5*max(size)) 2: set up data frame
    ## for by-group processing
    pivot.table <- group_by(HO2.df, direction)
    ## 3: calculate the summary metrics
    options(dplyr.width = Inf)  ## to display all columns
    HO2.count <- length(HO2.df$return)
    pivot.table <- summarise(pivot.table, 
        return.avg = mean(return), return.sd = sd(return), 
        quantile.5 = quantile(return, 
            0.05), quantile.95 = quantile(return, 
            0.95), percent = (length(return)/HO2.count) * 
            100)
    # Construct transpose of pivot table
    # with xtable()
    require(xtable)
    pivot.xtable <- xtable(t(pivot.table), 
        digits = 2, caption = HO2.caption, 
        align = rep("r", 4), table.placement = "V")
    HO2.caption <- "Heating Oil No. 2: 1986-2016"
    output.list <- list(table = pivot.table, 
        xtable = pivot.xtable, df = HO2.df)
    return(output.list)
}
```

Test HO2_movement() with data and display results in a table with 2 decimal places.

```{r}
knitr::kable(HO2_movement(file = "data/nyhh02.csv")$table, 
    digits = 2)
```

Morale: more work today (build the function) means less work tomorrow (write yet another report).

3. Suppose we wanted to simulate future movements in HO2 returns. What distribution might we use to run those scenarios? Here, let's use the MASS package's fitdistr() function to find the optimal fit of the HO2 data to a parametric distribution. We will use the gamma distribution to simulate future heating oil #2 price scenarios.

```{r}
library(MASS)
HO2.data <- HO2_movement(file = "data/nyhh02.csv", 
    caption = "HO2 NYH")$df
str(HO2.data)
fit.gamma.up <- fitdistr(HO2.data[HO2.data$direction == 
    "up", "return"], "gamma", hessian = TRUE)
fit.gamma.up
# fit.t.same <-
# fitdistr(HO2.data[HO2.data$direction
# == 'same', 'return'], 'gamma',
# hessian = TRUE) # a problem here is
# all observations = 0
fit.t.down <- fitdistr(HO2.data[HO2.data$direction == 
    "down", "return"], "t", hessian = TRUE)
fit.t.down
fit.gamma.down <- fitdistr(-HO2.data[HO2.data$direction == 
    "down", "return"], "gamma", hessian = TRUE)  # gamma distribution defined for data >= 0
fit.gamma.down
```

```{r}

```